Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2023-12-29 alignpair_letter.tex
COmmit completed first round of manuscript changes. I ended up touching just about everything to make it feel more like a research paper and less like an application note.

COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; c.f.][]{holmes2001evolutionary}. We call this transducer the COATi FST. Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics. COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
It also lets the user provide a codon substitution matrix. A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons. This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.

@@

In order to compare COATi against other aligners on realistic datasets, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels. Briefly, we downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
database \citep{ensembl_hubbard_2002}. After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with five alignment methods: COATi, Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011}, MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
\citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}. Next we separated sequence pairs into two sets: (1) 6048 sequence pairs for which at least one aligner added gaps, and (2) 7761 sequence pairs for which all five aligners did not introduce any gaps.

@@

COATi, using the codon-triplet-mg model, obtained better results compared to a wide variety of alignment strategies.
It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $1.3 \cdot 10^{-76}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
%
In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}). Furthermore, the estimated evolutionary divergence from the alignments retrieved by COATi is substantially less overestimated than other methods (Table \ref{table:comp}, Supplemental Materials Figure 8).

Clustal$\Omega$ generated alignments via amino acid translations and obtained the highest average alignment error while having difficulties retrieving positive selection.
MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
PRANK, using a codon model, had an average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.

@@

To test how well COATi performs when the roles of reference and low-quality sequence are reverted, I aligned the 7761 simulated alignments using gorilla as the reference. Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons in the gorilla sequence on the remaining alignments. While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence. Despite this limitation, I analyzed the 4003 alignments and compared the results with all methods, including COATi using the human sequence as the reference. The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the ancestor in comparison to the reverse (Supplementary Materials Tab. 6). However, the results for COATi continue to be a significant improvement over other aligners.

Despite human and gorilla sequences having a relatively short evolutionary distance, COATi showed a biologically significant improvement over other methods, with an average alignment error nine-fold smaller than the next best method.
COATi is an FST-based application that can calculate the optimal alignment between a pair of sequences in the presence of artifacts using a statistical model.
Using COATi will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.

Future work include extending the indel FST to combine a 3-mer gap model with a frameshift parameter and weighing each indel phase differently to reflect known selection on indel phases \citep{zhu2022profiling}.
We also plan on comparing the marginal and triplet models to evaluate the implications of the marginalization.

@@

This research was funded by NSF award DBI-1929850.

\TODO{Reviewers, Editor, and COmmittee members}

>>>>

COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; c.f.][]{holmes2001evolutionary}. We call this transducer the COATi FST. Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics. COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
It also lets the user provide a codon substitution matrix. A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[c.f.][]{ranwez_macse_2011,ranwez_macse_2018}. This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.

@@

In order to compare COATi against other aligners on realistic datasets, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels. Briefly, we downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
database \citep{ensembl_hubbard_2002}. After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with five alignment methods: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ v1.2.4 via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE v2.06, (4) the DNA model of MAFFT v7.505, and (5) the codon model of PRANK v.150803 \citep{clustal_omega_sievers_2011,ranwez_macse_2018,katoh2013mafft,prank_loytynoja_2014}. Next we separated sequence pairs into two sets: (1) 6048 sequence pairs for which at least one aligner added gaps, and (2) 7761 sequence pairs for which all five aligners did not introduce any gaps.

@@

COATi obtained better results compared to a wide variety of alignment strategies. It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $1.3 \cdot 10^{-76}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1). Notably, the average alignment error of the second best protocol was six times larger than COATi's. In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}). Furthermore, the estimated evolutionary divergence from the alignments retrieved by COATi was substantially less overestimated than other methods (Table \ref{table:comp}, Supplemental Materials Figure 8).

Aligning sequences using amino-acid translation via ClustalΩ obtained the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances. These results are not surprising because alignments in amino-acid space only permit phase-3 gaps and are sensitive to frameshift artifacts, which inflate divergence estimates and impact selection detection. Conversely, we obtained better results when aligning in nucleotide space using both amino-acid aware MACSE and amino-acid agnostic MAFFT. These results indicate that aligning protein-coding sequences in nucleotide space works reasonably well if the sequences come from closely related species, such as humans and gorillas, as it supports all three gap phases and is robust to frameshift artifacts. If the sequences were further apart, we predict that amino-acid agnostic approaches would begin to produce unreasonable alignments. And finally, our PRANK protocol used a codon substitution model which should have been an improvement over both amino-acid and nucleotide models; however, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts. In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three.

@@

COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences. To test how well COATi performs when the roles of sequences are reversed, we realigned the 7761 benchmark sequence alignments using gorilla as the reference. Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons or other artifacts in 3758 simulated gorilla sequences. While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence. Despite this limitation, we analyzed the 4003 alignments and compared the results across methods, including COATi using the human sequence as the reference. The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the reference in comparison to the reverse (Supplementary Materials Tab. 6). However, the results for COATi continue to be a significant improvement over other aligners.

COATi uses finite-state transducers to align pairs of sequences using a codon-aware statistical model. While COATi has multiple modes and models, here we have focused on evaluating the utility of COATi FST for estimating pairwise alignments. We have shown that COATi offers a biologically significant improvement over other methods when aligning pairs of sequences separated by short evolutionary distances in the presence of genomic artifacts. In other work, we have begun to explore application of COATi to more divergent sequences and the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.

COATi is under active development. We plan on extending it to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}. We plan on improving its multiple-sequence alignment and alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes. Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.

@@

This research was funded by NSF award DBI-1929850.
%
The authors would like to thank Profs.\ Marco Mangone, Ted Pavlic, Banu Oskan, Jay Taylor, and Jeremy Wideman for their helpful support on two separate PhD dissertation projects. The authors would also like to thank the associate editor and two reviewers for their helpful comments on earlier versions of this manuscript.
